---
title: "P8105 Homework 2"
author: "Chengshun Zhao"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Problem 1
```{r}
library(tidyverse)

subway_df = read_csv(
  "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"
)

```
```{r}
subway_df = subway_df |>
  janitor::clean_names() |>
  select(
    line, station_name,
    station_latitude, station_longitude,
    starts_with("route"),
    entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = entry == "YES"
  )
```
```{r}
dim(subway_df)

n_distinct_stations = subway_df |>
  distinct(line, station_name) |>
  nrow()

n_distinct_stations

n_ada_stations = subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

n_ada_stations

entry_proportion = subway_df |>
  filter(vending == "NO") |>
  summarize(proportion = mean(entry))

entry_proportion

unique(subway_df$vending)
```
Among station entrances/exits without vending, approximately 37.7% allow entry.
```{r}
route_df = subway_df |>
  mutate(across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )

a_train_stations = route_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_stations

a_train_ada = route_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

a_train_ada
```
The subway dataset contains 1868 observations and 19 variables. Variables include subway line and station names, longitude and latitude of stations, routes served, station entrance status, availability of vending machines, type of station entrance, and if the station is ADA compliant. I converted the entry variable from character values (“YES” or “NO”) to logical values, and select the variables to be included in the analysis. The dataset is not fully tidy, as the routes served are stored in multiple columns (route1 through route11). The routes should be stored as separate variables from the number and name of the routes.

There are 60 distinct stations that serve the A train. Of these stations, 17 are ADA compliant.

# Problem 2
```{r}
library(readxl)

excel_sheets("data/202610 Trash Wheel Collection Data.xlsx")

mr_trash = read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N1000"
) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

dim(mr_trash)
head(mr_trash)
```

```{r}
prof_trash = read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel",
  range = "A2:M1000"
) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Professor Trash Wheel"
  )

dim(prof_trash)
```
```{r}
gwynnda = read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynnda the Good Wheel of the W",
  range = "A2:M1000"
) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Gwynnda"
  )

dim(gwynnda)
```
```{r}
mr_trash = mr_trash |>
  mutate(
    year = as.numeric(year),
    across(plastic_bottles:wrappers, as.numeric)
  )

prof_trash = prof_trash |>
  mutate(
    year = as.numeric(year),
    across(plastic_bottles:wrappers, as.numeric)
  )

gwynnda = gwynnda |>
  mutate(
    year = as.numeric(year),
    across(plastic_bottles:wrappers, as.numeric)
  )

trash_data = bind_rows(
  mr_trash,
  prof_trash,
  gwynnda
)

dim(trash_data)
```
```{r}
prof_total_weight = trash_data |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(total_weight = sum(weight_tons, na.rm = TRUE))

prof_total_weight

gwynnda_june_2022 = trash_data |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(total_butts = sum(cigarette_butts, na.rm = TRUE))

gwynnda_june_2022
```
The merged Trash Wheel dataset comprises 1269 observations and 16 variables. The key variables in the dataset include the number of dumpsters, month, year, date, weight in tons, volume in cubic yards, plastic bottles, cigarette butts and the name of the Trash Wheel. According to the data collected by Professor Trash Wheel, a total of 214.96 tons of trash was collected. In June 2022, Gwynnda collected a total of 18,120 cigarette butts.
 
# Problem 3
```{r}
baseline_df = read_csv(
  "data/MCI_baseline.csv",
  skip = 1,
  na = "."
) |>
  janitor::clean_names()

head(baseline_df)
names(baseline_df)
dim(baseline_df)
```
```{r}
baseline_df = baseline_df |>
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  )

head(baseline_df)
```
```{r}
baseline_mci = baseline_df |>
  filter(!is.na(age_at_onset), age_at_onset <= current_age)

baseline_mci
```
```{r}
baseline_df = baseline_df |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

dim(baseline_df)
```
```{r}
baseline_summary = baseline_df |>
  summarize(
    participants = n(),
    developed_mci = sum(!is.na(age_at_onset)),
    mean_baseline_age = mean(current_age)
  )

baseline_summary
```

```{r}
female_apoe4 = baseline_df |>
  filter(sex == "Female") |>
  summarize(
    proportion = mean(apoe4 == "Carrier")
  )

female_apoe4
```
After importing and cleaning the baseline demographic data, sex and APOE4 carrier status were transformed to categorical variables. Four participants, who developed MCI at or before baseline and, therefore, did not meet the study inclusion criteria, were removed. The analysis was performed on the 479 participants. Of these participants, 93 developed MCI at some point during the follow-up. The average age at baseline was approximately 65.03 years. Of the women in the study, 30% were APOE4 carriers.

```{r}
amyloid_df = read_csv(
  "data/mci_amyloid.csv",
  skip = 1,
  na = "."
) |>
  janitor::clean_names()

head(amyloid_df)
names(amyloid_df)
dim(amyloid_df)
```
```{r}
amyloid_long = amyloid_df |>
  pivot_longer(
    baseline:time_8,
    names_to = "visit",
    values_to = "amyloid"
  )

head(amyloid_long)
dim(amyloid_long)

sum(is.na(amyloid_long$amyloid))

n_distinct(amyloid_long$study_id)
```
```{r}
baseline_only = baseline_df |>
  anti_join(amyloid_long, by = c("id" = "study_id")) |>
  distinct(id)

amyloid_only = amyloid_long |>
  anti_join(baseline_df, by = c("study_id" = "id")) |>
  distinct(study_id)

baseline_only
amyloid_only

combined_df = baseline_df |>
  inner_join(
    amyloid_long,
    by = c("id" = "study_id")
  )

dim(combined_df)
head(combined_df)
```
```{r}
write_csv(combined_df, "data/mci_amyloid_clean.csv")
```

The baseline dataset included 479 participants after removing 4 participants with MCI at baseline. Among the 479 participants, 93 developed MCI over the course of the study. The average age at baseline was 65.03 years. Thirty percent of the women were APOE4 carriers. The amyloid dataset contained 487 participants. This dataset was converted to long format and resulted in 5 observations per participant. There were 8 participants in the baseline dataset and 16 participants in the amyloid dataset only. Keeping only the participants in both datasets resulted in 471 participants, 2355 observations, and 8 variables.